---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r, echo = FALSE}
description <- readLines("DESCRIPTION")
version <- description[grepl(pattern = "^Version:", x = description)]
version <- substr(version, nchar("Version: ") + 1, nchar(version))
```

# heuristika `r version`

<!-- badges: start -->
<!-- badges: end -->

The goal of heuristika is to provide a computationally and outlier-robust implementation of a subset of exponential smoothing-based models that allows for manual overwrites.

Main characteristics of heuristika are:

- Lightweight with regard to dependencies
- Computational robustness
- Based on Exponential Smoothing models (additive trend, seasonality, error)
- Forecasts via sample path generation
- User-controllable robust handling of "outliers" as part of model fitting
- User-controllable initial states that could be shared across time series to share hierarchical information (possible due to standardization of time series)
- User-controllable parameter grid combined with custom penalized loss functions
- Choice of "likelihood function" for the error component

The implementation is based on a bunch of heuristics. Thus the name.

## Installation

You can install the development version of heuristika from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("timradtke/heuristika")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(heuristika)

y <- rnorm(55, mean = 25 + sinpi(1:55 / 6), sd = 0.5)

fitted_model <- heuristika::fit(y = y, m = 12, family = "norm")
```

```{r plot_fit, echo = FALSE}
plot(x = 1:67, y = c(y, rep(NA, 12)), pch = 19, 
     ylim = c(min(y)*0.95, max(y)*1.05),
     xlab = NA, ylab = NA)
lines(1:67, c(y, rep(NA, 12)))
lines(fitted_model$y_hat, col = rgb(red = 0, green = 0, blue = 1))
lines(fitted_model$x_cleaned, col = rgb(red = 0, green = 0, blue = 1))
```

```{r}
fc <- heuristika::forecast(
  object = fitted_model, 
  h = 12, 
  n = 10000
)
```

```{r plot_sample_paths}
plot(x = 1:67, y = c(y, rep(NA, 12)), pch = 19, 
     ylim = c(min(y)*0.95, max(y)*1.05),
     xlab = NA, ylab = NA)
lines(1:67, c(y, rep(NA, 12)))

for (j in 1:5) {
  lines(56:67, fc[, j], col = j + 1)
}
```

```{r plot_quantiles}
quantiles <- c(0.01, 0.05, 0.13, 0.5, 0.87, 0.95, 0.99)

plot(x = 1:67, y = c(y, rep(NA, 12)), pch = 19, 
     ylim = c(min(y)*0.95, max(y)*1.05),
     xlab = NA, ylab = NA)
lines(x = 1:67, y = c(y, rep(NA, 12)))
for (j in seq_along(quantiles)) {
  lines(
    x = 56:67, 
    y = apply(fc, quantile, MARGIN = 1, quantiles[j], na.rm = TRUE), 
    col = rgb(red = 0, green = 0, blue = 1, 
              alpha = 1 - abs(0.5 - quantiles[j])),
    lwd = 1 - abs(0.5 - quantiles[j])
    )
}
```

## AirPassengers

```{r airpassengers}
air_passengers <- log1p(as.numeric(AirPassengers))
ap_n <- length(air_passengers)

tictoc::tic()
ap_fit <- heuristika::fit(
  y = air_passengers[1:(ap_n-12)], m = 12, family = "norm"
)
tictoc::toc()

tictoc::tic()
ap_fc <- heuristika::forecast(object = ap_fit, h = 12, n = 10000)
tictoc::toc()
```

```{r airpassengers_plot, echo = FALSE}
plot(expm1(air_passengers), xlim = c(80, ap_n), ylim = c(0, 800), col = "white")
lines(expm1(ap_fit$y_hat), col = "blue")
for (j in seq_along(quantiles)) {
  lines(
    x = (ap_n-12+1):ap_n, 
    y = apply(expm1(ap_fc), quantile, MARGIN = 1, quantiles[j], na.rm = TRUE), 
    col = rgb(red = 0, green = 0, blue = 1, 
              alpha = 1 - abs(0.5 - quantiles[j])),
    lwd = 1 - abs(0.5 - quantiles[j])
  )
}
points(expm1(air_passengers), pch = 19, col = "black", cex = 0.5)
```

## deaths

```{r deaths}
deaths <- as.numeric(fdeaths)
d_n <- length(deaths)

tictoc::tic()
d_fit <- heuristika::fit(
  y = deaths[1:(d_n-12)], m = 12, family = "norm", 
)
tictoc::toc()

tictoc::tic()
d_fc <- heuristika::forecast(
  object = d_fit, 
  h = 12, 
  n = 10000,
  switch_to_cauchy_if_outliers = FALSE
)
tictoc::toc()
```

```{r deaths_plot, echo = FALSE}
plot(deaths, ylim = c(0, 1500), col = "white")
lines(d_fit$y_hat, col = "blue")
for (j in seq_along(quantiles)) {
  lines(
    x = (d_n-12+1):d_n, 
    y = apply(d_fc, quantile, MARGIN = 1, quantiles[j], na.rm = TRUE), 
    col = rgb(red = 0, green = 0, blue = 1, 
              alpha = 1 - abs(0.5 - quantiles[j])),
    lwd = 1 - abs(0.5 - quantiles[j])
  )
}
points(deaths, pch = 19, col = "black", cex = 0.5)
```

## References

Michael Bohlke-Schneider, Shubham Kapoor, Tim Januschowski (2020). *Resilient Neural Forecasting Systems*. https://www.amazon.science/publications/resilient-neural-forecasting-systems

Ruben Crevits and Christophe Croux (2017). *Forecasting using Robust Exponential Smoothing with Damped Trend and Seasonal Components*. KBI_1741, https://dx.doi.org/10.2139/ssrn.3068634

Rob J. Hyndman and George Athanasopoulos (2021). *Forecasting: Principles and Practice*. 3rd edition, OTexts: Melbourne, Australia. https://otexts.com/fpp3/

Rob J. Hyndman, Anne B. Koehler, Ralph D. Snyder, and Simone Grose (2002). *A State Space Framework for Automatic Forecasting using Exponential Smoothing Methods*. https://doi.org/10.1016/S0169-2070(01)00110-8

Rafael de Rezende, Katharina Egert, Ignacio Marin, Guilherme Thompson (2021). *A White-boxed ISSM Approach to Estimate Uncertainty Distributions of Walmart Sales*. https://arxiv.org/abs/2111.14721
